---
title: "Homework Template"
subtitle: "SOC 506: Quantitative Techniques in Sociology"
author: "[PUT YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
## Set knitr options here
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
knitr::opts_chunk$set(cache = TRUE)

```

```{r libraries, include=FALSE}
## load necessary libraryies her

## List of packages used
lop <- c("tidyverse","maps","classInt","XML","RCurl","rlist",'kableExtra')

## Check if installed
new.packages <- lop[!(lop %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

## Load packages
for(i in lop){library(package=i,character.only = TRUE) }

## Load non CRAN Packages
if(!'bbplot'%in%installed.packages()[,"Package"]){
  devtools::install_github('bbc/bbplot')
}

library('bbplot')
```


**Notes about this Template**

* This is an example RMarkdown doc to help prepare for the homeworks and labs coming up in the next couple weeks.
* Rmarkdown is very powerful.
* This demonstrates some of the neat things that can be done, you should look to play with it and learn more about it as the quarter progresses.



# Question 1

The New York Times maintains a set COVID-19 cases and deaths by state and county for the US. Artical can be found [here](https://www.nytimes.com/article/coronavirus-county-data-us.html?smid=ig-nytimes&utm_source=like2buy.curalate.com&crl8_id=1b624724-1c1a-4e0d-bb43-f4d32455ea31) and the github repo to access the data can be found [here](https://github.com/nytimes/covid-19-data).

Let's build a map of state level *cases* and county level *cases*.

# Solution 

## Visualization of COVID-19 {.tabset}

#### Data
```{r state_data,results='hide',message=FALSE}
## Load NYT COVID Data
covid19_state<-read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")
date<-'2020-03-25'
```

We will consider `r date` for this Homework template.

```{r}
## Select out 03/25 date
covid19_state_2020_03_25<-covid19_state %>% filter(date==date)

## load United States state map data
states <- map_data("state")

## rename state to region
names(covid19_state_2020_03_25)[2]<-"region"
## lowercase region
covid19_state_2020_03_25$region<-tolower(covid19_state_2020_03_25$region)

## merge COVID data with Map data
mergeStates <- inner_join(states, covid19_state_2020_03_25, by = "region")
mergeStates<-as_tibble(mergeStates)
```

### State 

```{r,out.width = '1200'}


classes <- classIntervals(covid19_state_2020_03_25$cases, n = 5, style = "jenks")

scale<-list(
  min = min(covid19_state_2020_03_25$cases),
  max = max(covid19_state_2020_03_25$cases),
  breaks = classes$brks
)


g <- ggplot() + geom_polygon(data=mergeStates, 
            aes(x=long, y=lat, group=group, fill = cases), 
              color="black", size = 0.2) + 
            scale_fill_continuous(name="COVID-19 Cases", low = "lightblue", 
            high = "darkblue",limits = c(scale$min,scale$max), breaks=scale$breaks, 
               na.value = "grey50",trans="log") +
           labs(title="COVID-19 Cases 03/25/2020")+
           theme(legend.position = "bottom",legend.key.width = unit(2, "cm"),legend.text = element_text(angle = 270, vjust=.5))
g
```


### County

```{r county_data,results='hide',message=FALSE}
## Load US County Data
data(county.fips)
## load United States state map data
counties<- map_data("county")
## set up fips file
fips.codes <- county.fips %>%
  # create two new variables, splitting "polyname" on the comma
  separate(polyname, c("region", "subregion"), ",")



## Load NYT COVID Data
covid19_county<-read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")
covid19_county$fips<-as.numeric(covid19_county$fips)

counties <- full_join(counties, fips.codes, by=c("region","subregion"))


```

```{r}
## Select out 03/25 date
covid19_county_2020_03_25<-covid19_county %>% filter(date==date)

## Combine 03/25 data with county data
counties<- full_join(counties, covid19_county_2020_03_25, by=c("fips"="fips"))

```


```{r,out.width = '1200'}

classes <- classIntervals(covid19_county_2020_03_25$cases, n = 5, style = "jenks")

scale<-list(
  min = min(covid19_county_2020_03_25$cases),
  max = max(covid19_county_2020_03_25$cases),
  breaks = classes$brks
)


g <- ggplot() + geom_polygon(data=counties, 
            aes(x=long, y=lat, group=group, fill = cases), 
              color="black", size = 0.2) + 
            scale_fill_continuous(name="COVID-19 Cases", low = "lightblue", 
            high = "darkblue",limits = c(scale$min,scale$max), breaks=scale$breaks, 
               na.value = "grey50",trans="log") +
           labs(title="COVID-19 Cases by County, 03/25/2020")+ 
           theme(legend.position = "bottom",legend.key.width = unit(2, "cm"),   
                 legend.text = element_text(angle = 270, vjust=.5))
           
           
g
```

# Question 2

# Solution

## State Curves {.tabset}

### Washington
```{r}

st<-covid19_state %>% filter(state=='Washington')
line <- ggplot(st, aes(x = date, y = cases)) +
  geom_line(colour = "#1380A1", size = 1) +
  geom_hline(yintercept = 0, size = 1, colour="#333333") +
  bbc_style() +
  labs(title="Washington COVID-19 Cases",
       subtitle = "NYT Data")
line
```

### New York
```{r}

st<-covid19_state %>% filter(state=='New York')
line <- ggplot(st, aes(x = date, y = cases)) +
  geom_line(colour = "#1380A1", size = 1) +
  geom_hline(yintercept = 0, size = 1, colour="#333333") +
  bbc_style() +
  labs(title="New York COVID-19 Cases",
       subtitle = "NYT Data")
line
```

### California
```{r}
st<-covid19_state %>% filter(state=='California')
line <- ggplot(st, aes(x = date, y = cases)) +
  geom_line(colour = "#1380A1", size = 1) +
  geom_hline(yintercept = 0, size = 1, colour="#333333") +
  bbc_style() +
  labs(title="California COVID-19 Cases",
       subtitle = "NYT Data")
line

```

# Question 3

# Solution

## Washington State {.tabset}

```{r}
## use google cache verision to get around java script issues
## https://www.doh.wa.gov/Emergencies/Coronavirus
theurl <- getURL("https://webcache.googleusercontent.com/search?q=cache:PjWUDebdvq4J:https://www.doh.wa.gov/emergencies/coronavirus+&cd=1&hl=en&ct=clnk&gl=us",.opts = list(ssl.verifypeer = FALSE) )
tables <- readHTMLTable(theurl)
tables <- list.clean(tables, fun = is.null, recursive = FALSE)
```

### Table: Confirmed Cases and Deaths by County
```{r}
#for(i in 1:length(tables)){
  i = 1
  kable(tables[[i]]) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "500px", height = "200px")
#}
```

### Table: Count of Indivuals Tested
```{r}
#for(i in 1:length(tables)){
  i = 2
  kable(tables[[i]]) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "500px", height = "200px")
#}
```


### Table: Confirmed Cases / Deaths by Age
```{r}
#for(i in 1:length(tables)){
  i = 3
  kable(tables[[i]]) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "500px", height = "200px")
#}
```

### Table: Confirmed Cases / Deaths by Gender
```{r}
#for(i in 1:length(tables)){
  i = 4
  kable(tables[[i]]) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "500px", height = "200px")
#}
```

### Confirmed Cases


```{r}
smart.round <- function(x) {
  y <- floor(x)
  indices <- tail(order(x-y), round(sum(x)) - sum(y))
  y[indices] <- y[indices] + 1
  y
}


## Assume 80+ is 12%
age_count_tmp<-3723*c(.02,.10,.15,.14,.18,.16,.13,.12)
age_count<-smart.round(age_count_tmp)

age_group<-tables[[3]]
names(age_group)<-gsub(" ","_",names(age_group))
age_group$percent_of_cases_numeric<-c(.02,.10,.15,.14,.18,.16,.13,.13)
age_group$percent_of_death_numeric<-c(0,0,0,.01,.06,.11,.28,.54)
age_group<-cbind(age_group,age_count)

ggplot(age_group, aes(Age_Group, percent_of_cases_numeric )) + 
  geom_bar(stat = 'identity',colour = "white", fill = "#1380A1")+
  bbc_style()+
  labs(title = "Washington COVID Cases",
       subtitle = "Age Distribution of Confirmed Cases, 03/27/20")
```

### Confirmed Deaths

```{r}
ggplot(age_group, aes(Age_Group, percent_of_death_numeric)) + 
  geom_bar(stat = 'identity',colour = "white", fill = "#1380A1")+
  bbc_style()+
  labs(title = "Washington COVID Cases",
       subtitle = "Age Distribution of Deaths, 03/27/20")


```

