---
title: "Week 1 Lab"
subtitle: "SOC 506: Quantitative Techniques in Sociology"
author: "Zack W. Almquist, University of Washington"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
## Set knitr options here
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA, 
                      cache = TRUE,
                      warning = FALSE,
                      message = FALSE, 
                      cache.lazy = FALSE
                      )

```

```{r libraries, include=FALSE}
## load necessary libraryies her

## List of packages used
lop <- c("tidyverse","maps","classInt","XML","RCurl","rlist",'kableExtra','alr3')
git_lop<-list(
  author = c('bbc','kjhealy'),
  package = c('bbplot','gssr')
)
## Check if installed
new.packages <- lop[!(lop %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

## Load packages
for(i in lop){library(package=i,character.only = TRUE) }

## check if installed
new.packages <- lapply(git_lop,function(x){x[!(git_lop$package %in% installed.packages()[,"Package"])]})

if(length(new.packages$package)>0){
  for(i in 1:length(new.packages$package)){
    devtools::install_github(paste(new.packages$author[[i]],new.packages$package[[i]],sep='/'))
  }
}


## Load git libraries
for(i in git_lop$package){library(package=i,character.only = TRUE) }

```
# Useful R Guides

* [Awesome Tables in HTML](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)
* [R for Data Science](https://r4ds.had.co.nz/) [Tidyverse Introduction]
  + [Tibbles data in R](https://r4ds.had.co.nz/tibbles.html)
  + [Tiddy data](https://r4ds.had.co.nz/tidy-data.html)
  + [Work flow basics](https://r4ds.had.co.nz/workflow-basics.html)
  + [RMarkdown](https://r4ds.had.co.nz/r-markdown.html)
* [RMarkdown Vignette for gssr package](https://github.com/kjhealy/gssr/blob/master/vignettes/overview.Rmd)

# Data and data management in R: A primer

In this section of the lab we are going to use data from the [General Social Survey](https://gss.norc.org/) made available in R throught the [gssr](https://kjhealy.github.io/gssr/) package.

## Basics of R, matrices and Data

For this discussion I have included [Charlie Geyer's](http://users.stat.umn.edu/~geyer/) R notes from his introduction to Computational Statistics course below. These notes are excellent and we will review them in course briefly.

* [R Basics](basic.html)
* [Arrays](array.html)
* [Data Basics](data.html)

## GSS {.tabset}

### GSS Code Book (10 entries)
```{r}
data(gss_doc)
  kable(gss_doc[1:10,]) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "900px", height = "400px")
```

### GSS Data (10 entries)

```{r}
data(gss_all)
  kable(gss_all[1:10,]) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "900px", height = "400px")
```

## GSS Tabulated Data GSS {.tabset}

### Race
```{r}
  race<- gss_doc %>% filter(id == "race") %>%
  select(marginals) %>% 
  unnest(cols = c(marginals))
  kable(race) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) #%>%
  #scroll_box(width = "900px", height = "400px")
```


### Sex (really gender)
```{r}
  sex<- gss_doc %>% filter(id == "sex") %>%
  select(marginals) %>% 
  unnest(cols = c(marginals))

  kable(sex) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) #%>%
  #scroll_box(width = "900px", height = "400px")
```

### Race and Sex

```{r}
  sex_race<- gss_get_marginals(varnames = c("race", "sex"))

  kable(sex_race) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) #%>%
  #scroll_box(width = "900px", height = "400px")
```


# Descriptive Statistics in R

## Base R

`summerize`, `table`, `quantile`, `mean`, `median`, `mode`, `sum`

```{r}
age<-gss_all %>% filter(year==max(year))%>% select(age)
head(age)
```

```{r}
#general summary statistics
summary(age[[1]])
```

```{r}
## Frequency table
tab<-data.frame(table(age,useNA = 'always'))
head(tab,10)

## Quartiles, general quantile function
quantile(age[[1]],probs=seq(0, 1, 0.25), na.rm = TRUE)

## mean with na's removed
mean(age[[1]],na.rm=TRUE)

## median with na's removed
median(age[[1]],na.rm=TRUE)

## mode function gives type
mode(age[[1]]) ##Ooops!

## We will have to write our own statistical mode function!
stat_mode<-function(x){
  tab<-as.data.frame(table(x))
  index<-which.max(tab$Freq)
  as.numeric(as.character(tab[index,1]))
}

##stat mode
stat_mode(age[[1]])

## sum with na's removed
sum(age[[1]],na.rm=TRUE)
```

## Tidyverse

```{r}
gss_all %>% 
  filter(year==max(year))%>% 
  select(age) %>% 
  summarise(mean = mean(age,na.rm=TRUE),n = n()) %>%
  t()

gss_all %>% 
  filter(year==max(year))%>% 
  select(age) %>% 
  summarise(mean = mean(age,na.rm=TRUE), 
            variance = var(age,na.rm=TRUE),n = n()) %>%
  t()

gss_all %>% 
  filter(year==max(year))%>% 
  select(age,childs) %>% summarise_all(list(mean=mean,median=median,var=var,sd=sd),na.rm=TRUE)%>%
  t()


gss_all %>% 
  filter(year%in%c(2016,2018))%>%
  group_by(year,age,sex) %>% 
  summarize(n())

```

# Statistical Tests in R

## permutation test {.tabset}

* [Permutation Test Notes form our Friends in Bio-Stats](http://faculty.washington.edu/kenrice/sisg/SISG-08-06.pdf)

> To estimate the sampling distribution of the test statistic we need many samples generated under the strong null hypothesis.

> If the null hypothesis is true, changing the exposure would have
no effect on the outcome. By randomly shuffling the exposures
we can make up as many data sets as we like.

>If the null hypothesis is true the shuffled data sets should look
like the real data, otherwise they should look different from the
real data.

> The ranking of the real test statistic among the shuffled test
statistics gives a p-value

### data

```{r}
## make up some ‘true’ data
carrier<-rep(c(0,1), c(100,200))
null.y<-rnorm(300)
alt.y<-rnorm(300, mean=carrier/2)

data.frame(null.y,alt.y,carrier)[sample(1:300,10),]
```

### t-test

```{r}
t.test(null.y~carrier, var.equal=TRUE)
t.test(alt.y~carrier, var.equal=TRUE)
```

### mean test

```{r}
null.diff<-mean(null.y[carrier==1])-mean(null.y[carrier==0])
alt.diff<-mean(alt.y[carrier==1])-mean(alt.y[carrier==0])
one.test <- function(x,y) {
xstar<-sample(x)
mean(y[xstar==1])-mean(y[xstar==0])
}
many.truenull <- replicate(1000, one.test(carrier, null.y))
many.falsenull <- replicate(1000, one.test(carrier, alt.y))

cat('Null Difference\n')
null.diff
cat(sum(abs(many.truenull)>abs(null.diff)),' out of 1000 replicates\n')
cat('\nAlt Difference\n')
cat(sum(abs(many.falsenull)>abs(alt.diff)),' out of 1000 replicates\n')


```

### mean test: histogram

```{r}
hist(many.truenull)
abline(v=null.diff, lwd=2, col="purple")
mean(abs(many.truenull) > abs(null.diff))
```
`r sum(abs(many.truenull) > abs(null.diff))` shuffled differences exceeed true difference: p = `r abs(null.diff)`


```{r}
hist(many.falsenull,xlim=c(min(many.falsenull),max(alt.diff)))
abline(v=alt.diff, lwd=2, col="purple")
mean(abs(many.falsenull) > abs(alt.diff))
```

`r sum(abs(many.falsenull) > abs(alt.diff))` shuffled differences exceeed true difference: p = `r abs(null.diff)`

# Linear Regression in R

* [Notes on SLR](SLR_Slides.pdf)


