---
title       : SOC 506 - R Lecture 1
subtitle    : Basics of R
author      : Zack W Almquist
job         : University of Washington
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax, quiz, bootstrap, interactive] # {mathjax, quiz, bootstrap}
ext_widgets : {rCharts: [libraries/nvd3, libraries/leaflet, libraries/dygraphs]}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
logo        : W-Logo_RegistrationMark_Purple_26851.png
biglogo     : Signature_Left_Purple_RGB.png
assets      : {assets: ../../assets}
--- .class #id

<style type="text/css">
body {background:grey transparent;
}
</style>

## Course overview

In this lecture we will cover
- R
- Software, IDEs and other necessary details
- Some key aspects of R as Programing Language

--- .segue bg:grey

# First things first...

--- .class #id

## Why R?

- [R](https://www.r-project.org/)


--- .class #id

## Why RStudio?

- [Rstudio](https://www.rstudio.com/)

--- .class #id

## tidyverse

- [tidyverse](https://www.tidyverse.org/)

--- .class #id

## Installing packages

- CRAN Install
```{r,eval=FALSE}
install.packages("here")
```
- Github Install

```{r,eval=FALSE}
devtools::install_github("mages/googleVis")
```

--- .class #id


## RMarkdown and Homeworks, Labs and Final Project

- All Labs, HWs and Final Project will be written in RMarkdown and compiled to HTML
- This makes a transparent document that includes all necessary R code
- THis is part of the open and reproducible science movement.

--- .class #id

## Review of template

- Next session will cover some basics of RMarkdown.

--- .class #id

### Header 

```
HEADER HERE
```

--- .class #id

## R code

- You can insert code by surrounding it with \`\`\`{r}, followed by \`\`\` (three more ticks)...

\`\`\`{r}  
code here  
\`\`\`  

- Here is an example of some simple math

```{r}
2 + 2
```

--- .class #id

## Overview of R basics

### Assignment

```{r}
foo <- 2
foo
foo = 3
foo
```

--- .class #id

## Overview of R basics

### Assignment

```{r}
4 -> foo
foo
foo <<- 5
foo
6 ->> foo
foo
```

--- .class #id

## Overview of R basics

### Assignment

```{r}
assign("foo", 7)
foo
```

--- .class #id

## Overview of R basics

### Caclulator

```{r}
6*8
4^3
exp(1)   # exp() is the exponential function
```

--- .class #id

## Overview of R basics

### Caclulator

```{r}
pi     # the constant 3.14159265...
sin(0)
log(5) # unless you specify the base, R will assume base e
log(5, base=10)  # base 10
```
```

--- .class #id

## Overview of R basics

### Functions

```{r}
ralph <- function(x) {
    x <- 2
    return(x)
}
x <- 1
ralph(x)
x
```

--- .class #id

## Overview of R basics

### for loops

```{r}
 for( i in 1:5 ){
        print(i)
    }
```

--- .class #id


## Overview of R basics

### data types: vectors

```{r}
one <- 1
one
is.vector(one)
length(one)
```

--- .class #id

## Overview of R basics

### data types: vectors

```{r}
1:40
```

--- .class #id

## Overview of R basics

### data types: vectors

```{r}
one <- 1
typeof(one)
mode(one)
storage.mode(one)
class(one)
```

--- .class #id

## Overview of R basics

### data types: vectors

```{r}
length(LETTERS)
head(LETTERS)
tail(LETTERS)
length(colors())
head(colors())
tail(colors())
length(1:20)
length(double())
```

--- .class #id

## Overview of R basics

### data types: vectors

```{r}
1:5 + 6:10
1:5 * 6:10
1:5 ^ 6:10
```

--- .class #id

## Overview of R basics

### data types

```{r}
as.character(1:5)
as.numeric(letters[1:5])
as.logical(0:5)
```

--- .class #id

## Overview of R basics

### data.frames

```{r}
Grades <- data.frame(
  Name  = c('Bob','Jeff','Mary','Valerie'), 	
  Exam.1 = c(90, 75, 92, 85),
  Exam.2 = c(87, 71, 95, 81)
)

```

--- .class #id


## Overview of R basics

### data.frames

R allows two differnt was to access elements of the `data.frame`. First is a matrix-like notation for accessing particular values. 

  Format  |   Result
--------- | --------------------------------------
`[a,b]`   |  Element in row `a` and column `b`
`[a,]`    |  All of row `a`
`[,b]`    |  All of column `b`

```{r}
Grades[, 2]       # print out all of column 2 
Grades$Name       # The $-sign means to reference a column by its label
```

--- .class #id

## Overview of R basics

### ifelse statement

```{r}
x <- rnorm(6)
x
ifelse(x < 0, x - 3, x + 3)
x + 3 * sign(x)
```

--- .class #id

## Overview of R basics

### apply functions

```{r}
# Create the matrix
m<-matrix(c(seq(from=-98,to=100,by=2)),nrow=10,ncol=10)

# Return the product of each of the rows
apply(m,1,prod)

# Return the sum of each of the columns
apply(m,2,sum)

```

--- .class #id

## Overview of R basics

### apply functions

```{r}

# Return a new matrix whose entries are those of 'm' modulo 10
apply(m,c(1,2),function(x) x%%10) 
```

--- .class #id



